Problem Set 5
========================================================

This assignment is the first part of problem set 5 of the Udacity course *Data Analysis with R* lesson 5: *Explore Many Variables*. This part of the problem set consists in a set of questions/tasks answered below.

### Diamonds
Load the diamonds dataset

```{r}
suppressMessages(library(ggplot2))
data(diamonds)
str(diamonds)
```

***

### Price Histograms with Facet and Color
Create a histogram of diamond prices. Facet the histogram by diamond color and use cut to color the histogram bars.
```{r fig.width=10, fig.height=10}
ggplot(aes(x = price), data = diamonds) +
  geom_histogram(binwidth = 200, aes(fill = cut)) +
  facet_wrap(~color)

## scales = free_y
ggplot(aes(x = price), data = diamonds) +
  geom_histogram(binwidth = 200, aes(fill = cut)) +
  facet_wrap(~color, scales = "free_y")
```

***

### Price vs. Table Colored by Cut
Create a scatterplot of diamond price vs. table (width of top of diamond relative to widest point) and color the points by the cut of the diamond.
```{r}
ggplot(aes(x = table, y = price), data = diamonds) +
  geom_point(aes(color = cut)) +
  coord_cartesian(xlim = c(50,80))
```

#### What is the typical table range for the majority of diamonds of **ideal** cut?
53-57

#### What is the typical table range for the majority of diamonds of **premium** cut?
58-63

***

### Price vs. Volume and Diamond Clarity
Create a scatterplot of diamond price vs. volume (x * y * z) and color the points by the clarity of diamonds. Use scale on the y-axis to take the log10 of price. You should also omit the top 1% of diamond volumes from the plot.  
Note: Volume is a very rough approximation of a diamond's actual volume.

```{r}
diamonds$volume <- diamonds$x * diamonds$y * diamonds$z
ggplot(aes(x = volume, y = price), data = subset(diamonds, volume > 0 & volume < quantile(diamonds$volume, .99))) +
  geom_point(aes(color = clarity)) +
  scale_y_log10()
```

***

###  Proportion of Friendships Initiated

#### Load Pseudo Facebook Friends Dataset
```{r}
pf <- read.csv('pseudo_facebook.tsv', sep = '\t')
```

#### Add new var, proportion of friendships initiated
```{r}
pf$prop_initiated <- pf$friendships_initiated / pf$friend_count
```

### prop_initiated vs. tenure
Create a line graph of the median proportion of friendships initiated ('*prop_initiated*') vs. *tenure* and color the line segment by *year_joined.bucket*.

From lesson 5, add var *year_joined* and *year_joined.bucket* to the dataset. The vars represent the year the user joined Facebook and the interval of years, respectively.
```{r}
pf$year_joined <- 2014 - ceiling(pf$tenure / 365)
table(pf$year_joined)

pf$year_joined.bucket <- cut(pf$year_joined, breaks=c(2004,2009,2011,2012,2014))
table(pf$year_joined.bucket)

ggplot(aes(x = tenure, y = prop_initiated), data = subset(pf, !is.na(prop_initiated))) +
  geom_line(stat = "summary", fun.y = median, aes(color = year_joined.bucket))
```

***

### Smoothing prop_initiated vs. tenure
You can use larger bins for tenure or add a smoother to the plot.
```{r}
## larger bins
ggplot(aes(x = 30 * round(tenure / 30), y = prop_initiated), data = subset(pf, !is.na(prop_initiated))) +
  geom_line(stat = "summary", fun.y = median, aes(color = year_joined.bucket))

## smoother
ggplot(aes(x = tenure, y = prop_initiated), data = subset(pf, !is.na(prop_initiated))) +
  geom_smooth(aes(color = year_joined.bucket))
```

#### Which group has the highest proportion of friendships initiated ?
Group [2012-2014]

#### For this group, what is the average(mean) proportion of friendships initiated?
0.67
```{r}
suppressMessages(library(dplyr))
pf %>%
  filter(!is.na(prop_initiated)) %>%
  group_by(year_joined.bucket) %>%
  summarise(mean = mean(prop_initiated))
```


#### Why do you think this happens?
I think users that have just joined Facebook are going to be more interested in looking for friends for several reasons: 

1. Fewer friends, since they have have joined just recently.
2. People they already know, already on Facebook.
3. New users are more engaged

***

### Price/Carat Binned, Faceted, & Colored
Back to the diamonds dataset. Create a scatter plot of the price/carat ratio of diamonds. The variable x should be assigned to cut. The points should be colored by diamond color, and the plot should be faceted by clarity.
```{r fig.width=10, fig.height=10}
ggplot(aes(x = cut, y = price/carat), data = diamonds) +
  geom_point(aes(color = color), position= position_jitter()) +
  facet_wrap(~clarity)
```

***
Sara Carvalho  
10-03-2015