Problem Set 3 
========================================================

This assignment is the first part of problem set 3 of the Udacity course *Data Analysis with R* lesson 3: *Explore One Variable*. This part of the problem set consists in a set of questions/tasks answered below.

### Diamonds
Load the diamonds dataset.

```{r}
library(ggplot2)
data(diamonds)
?diamonds

str(diamonds)
```


#### How many observations are in the dataset?
53940

#### How many variables ?
10

#### How many ordered factors are in the dataset?
3

#### What letter represents the best color for a diamond?
D 

***

### Price Histogram
Create a histogram of the price of all the diamonds in the diamond data set.

```{r fig.width= 10}
qplot(x = price, data = diamonds, binwidth = 500,
      color = I('black'), fill = I('light blue')) + 
  scale_x_continuous(breaks = seq(0,19000, 1000))
```

#### Describe the shape and center of the price distribution.
The distribution is right-skewed and presents a long tail. In such a distribution, the *median* is often a better representation of the center price of the diamonds then the *mean*. 

```{r}
summary(diamonds$price)
```

***

### Diamonds Count

#### How many diamonds cost less than $500?
```{r}
dim(subset(diamonds, price < 500))[1]
```

#### How many diamonds cost less than $250?
```{r}
dim(subset(diamonds, price < 250))[1]
```

#### How many diamonds cost $15,000 or more?
```{r}
dim(subset(diamonds, price >= 15000))[1]
```

***

### Cheaper Diamonds
Explore the largest peak in the price histogram you created earlier.
Try limiting the x-axis, altering the bin width, and setting different breaks on the x-axis.

```{r}
qplot(x = price, data = diamonds, binwidth = 25,
      color = I('black'), fill = I('grey')) +
  scale_x_continuous(breaks=seq(0,5000, 500), limits=c(0,5000))
```

The largest peak seems to be in the [300, 1500] interval.

```{r fig.width= 9, fig.height=12}
library(gridExtra)

p1 <- qplot(x = price, data = diamonds, binwidth = 5,
      color = I('black'), fill = I('grey'),
      main = "Price Histogram (binwidth = 5)") +
  scale_x_continuous(breaks=seq(325,1500, 200)) + 
  coord_cartesian(xlim = c(200,1500))

p2 <- qplot(x = price, data = diamonds, binwidth = 25,
      color = I('black'), fill = I('grey'),
      main = "Price Histogram (binwidth = 25)") +
  scale_x_continuous(breaks=seq(325,1500, 200)) + 
  coord_cartesian(xlim = c(200,1500))

p3 <- qplot(x = price, data = diamonds, binwidth = 50,
      color = I('black'), fill = I('grey'),
      main = "Price Histogram (binwidth = 50)") +
  scale_x_continuous(breaks=seq(325,1500, 200)) + 
  coord_cartesian(xlim = c(200,1500))

grid.arrange(p1,p2,p3)
```

Smaller bin-widths of 5, for example, result in a very "noisy" histogram that is difficult to read. 
Bin-widths of 25 and 50 are a better choice for a smoother and easy to read histogram.

```{r}
summary(subset(diamonds$price, diamonds$price > 325 & diamonds$price <=1500))
```

No diamonds cost 1500.  

***

### Price by Cut

#### Histograms

```{r}
qplot(x = price, data = diamonds, binwidth = 50)+
  facet_wrap(~cut) +
  coord_cartesian(xlim = c(300, 2000))
```

#### Which cut has the highest priced diamond ?
Premium
```{r}
by(diamonds$price, diamonds$cut, max)
```

#### Which cut has the lowest priced diamond ? 
Premium, Ideal
```{r}
by(diamonds$price, diamonds$cut, min)
```

#### Which cut has the lowest median price ? 
Ideal
```{r}
by(diamonds$price, diamonds$cut, median)
```

Note: the summary presents different results for some of these values. Example: max.
```{r}
by(diamonds$price, diamonds$cut, summary)
```

***

###  Scales and Multiple Histograms

Looking at the summary, the distributions should be somewhat similar, but the previous histograms created don't show that.  
Make scales free for histograms with *facet_warp*.

```{r}
qplot(x = price, data = diamonds, binwidth=250) + 
  facet_wrap(~cut, scales = 'free')
```

***

### Price per Carat by Cut
Create a histogram of price per carat and facet it by cut. Adjust the bin width and transform the scale of the x-axis using log10.

```{r}
qplot(x = price/carat, data = diamonds, binwidth = .05,
      fill=..count..) +
  facet_wrap(~cut, scales='free_y') +
  scale_x_log10()
```

***

### Price Box Plots
Investigate the price of diamonds using box plots, numerical summaries, and one of the following categorical variables: cut, clarity, or color. 

#### Color
```{r}
by(diamonds$price, diamonds$color, summary)

qplot(x = color, y = price, data = diamonds,
      geom = 'boxplot')
```

The prices seem to be higher for the J (worst) color. =/  
Perhaps a price/carat analysis would be more accurate.

***

### Interquartile Range - IQR

```{r}
by(diamonds$price, diamonds$color, summary)
```

#### What is the price range for the middle 50% of diamonds with color D ?
911 -  4214

#### What is the price range for the middle 50% of diamonds with color J ?
1860 - 7695

#### What is the IQR for diamonds with the best color?
```{r}
IQR(subset(diamonds, color == 'D')$price)
```

#### What is the IQR for diamonds with the worst color?
```{r}
IQR(subset(diamonds, color == 'J')$price)
```

***

### Price per Carat Box Plots by Color
```{r}
by(diamonds$price/diamonds$carat, diamonds$color, summary)

qplot(x = color, y = price/carat, data = diamonds,
      geom = 'boxplot') + 
  coord_cartesian(ylim=c(0,10000))

qplot(x = color, y = price/carat, data = diamonds,
      geom = 'boxplot')
```

The prices/carat don't show a significant difference by color. The better colors have, however, more and higher outliers.

#### Subset data for the middle 50% diamonds carats:
```{r}
summary(diamonds$carat)
subset.diamonds <- subset(diamonds, carat > 0.4 & carat < 1.05)

qplot(x = color, y = price/carat, data = subset.diamonds,
    geom = 'boxplot') + 
  coord_cartesian(ylim=c(1500,5000))
```

Looking only at the diamonds in the middle 50% carats, the difference in price by color is more visible.

***

### Carat Frequency Polygon
Investigate the weight of the diamonds (carat) using a frequency polygon. Use different bin widths to see how the frequency polygon changes.  
What carat size has a count greater then 2000 ?

```{r fig.width=10}
qplot(x = carat, data = diamonds, geom = 'freqpoly', binwidth = 0.1,
      main = "Carat Frequency polygon ( binwidth = 0.1)") +
  scale_x_continuous(breaks = seq(0,5,.1)) +
geom_abline(intercept = 2000, slope = 0, color = 'red', size = 1, lty = 'dashed')

qplot(x = carat, data = diamonds, geom = 'freqpoly', binwidth = 0.01,
      main = "Carat Frequency polygon ( binwidth = 0.01)") +
  scale_x_continuous(breaks = seq(1,1.1,.01)) +
  coord_cartesian(xlim=c(1,1.1))+
geom_abline(intercept = 2000, slope = 0, color = 'red', size = 1, lty = 'dashed')
```

#### What carat size has a count greater then 2000 ?
0.3, 1.01 (in exercise)

***

Sara Carvalho  
19-02-2015