Problem Set 3 - Gapminder
========================================================

## Introduction

This assignment is part of problem set 3 of the *Data Analysis with R* Udacity course's lesson 3: *Explore One Variable*. The task is to choose one of the datasets available in the Gapminder website and, making use of the techniques from lesson 3, create 2-5 plot to study the data. The Gapminder website contains over 500 data sets with information about the world's population.

### The Dataset

The dataset chosen from Gapminder is a subset of a larger dataset of *the average number of years of school attended by all people in the age and gender group specified, including primary, secondary and tertiary education*. The subset concerned only women between the age of 25-34. The source of the data is the Institute for Health Metrics and Evaluation and it was uploaded in 2010.

[Gapminder - Mean years in school - women 25-34](https://spreadsheets.google.com/spreadsheet/pub?key=0ArfEDsV3bBwCdC1MYzAtY2xPQ2xOR1lMeGhYSWlpR0E&gid=0)

***

## Dataset Investigation and Observations
This chapter shows the tools and code used in the investigation of the dataset, as well as the observations and notes taken along the study.

### Data Wrangling and First Notes

```{r tidy=TRUE}
school.years <- read.csv('Years.in.school.women.25-34.csv', header = TRUE, check.names = FALSE)
str(school.years)
```

This dataset needs some changes before it can be used properly. The first thing that I notice is that number of variables is really high, but then we learn that this is due to each year being considered as a different variable. We can collapse all this columns into one variable *year*, for which there is a value (the mean number of years).

```{r message=FALSE, }
library(tidyr)
library(dplyr)

names(school.years)[1] <- "country"
gather.school <- gather(data = school.years, key = "year", value ="mean.num.years", 2:length(school.years))
head(gather.school)
any(is.na(gather.school))
```

```{r}
str(gather.school)
```

Now the dataset is ready to use. We have 3 variables: *country*, *year*, *mean.num.years* (the mean number of years spend in school). Also, we can see that we have a value for every key, which is nice.

We can start by looking at the highest and lowest values.

#### 1. Countries with the highest values
```{r}
gather.school %>%
  arrange(desc(mean.num.years), desc(year)) %>%
  slice(1:5)
```


#### 2. Countries with the lowest values
```{r}
gather.school %>%
  arrange(mean.num.years, year) %>%
  slice(1:5)
```

***

### Distant vs Recent Years

We can observe that the highest and lowest values for the mean number of years spent in school appear also associated with the highest and lowest values for the *year* variable. So now, let's take a look at the values within the most recent and most distant years.

#### 1. Countries with the highest value in 1970
```{r}
gather.school %>% 
  filter(year == 1970) %>% 
  arrange(desc(mean.num.years)) %>% 
  slice(1:5)  %>% 
  select(country, mean.num.years)
```

#### 2. Countries with the lowest value in 1970
```{r}
gather.school %>% 
  filter(year == 1970) %>% 
  arrange(mean.num.years) %>% 
  slice(1:5)  %>% 
  select(country, mean.num.years)
```

#### 3. Countries with the highest value in 2009
```{r}
gather.school %>% 
  filter(year == 2009) %>% 
  arrange(desc(mean.num.years)) %>% 
  slice(1:5)  %>% 
  select(country, mean.num.years)
```

#### 4. Countries with the lowest value in 2009
```{r}
gather.school %>% 
  filter(year == 2009) %>% 
  arrange(mean.num.years) %>% 
  slice(1:5)  %>% 
  select(country, mean.num.years)
```

***

### Mean Number of Years In School by Year

#### Basic statistics for the values by decade
First we take a look at the basic statistics for the values per year. However, because there are values from 40 different years, let's take a look at a smaller subset of those values: per decade. 
```{r}
gather.school %>%
  group_by(year) %>%
  summarise(min = min(mean.num.years),
            first.q = quantile(mean.num.years, probs = 0.25),
            median = median(mean.num.years),
            mean = mean(mean.num.years), 
            third.q = quantile(mean.num.years, probs = 0.75),
            max = max(mean.num.years)) %>%
  filter(year %in% c(1970,1980,1990,2000,2009))
```
We can see that there is an increase in nearly every statistic for the values from one decade to the other.

#### Box plots and frequency polygon
In the next plots take a look at the distribution of the values (mean number of years in school) in the different years. 
The first plots are box plots, which provide a general view of the distribution of the dataset and, also, the evolution of its basic statistical values . Then, we look at the same values in 10 years intervals. Finally, we will use a frequency polygon to analyze the number of countries that fall into each year interval of mean years in school, and their evolution throughout the decades.

```{r fig.width=10}
library(ggplot2)

qplot(data = gather.school, x = year, y = mean.num.years, 
      geom = 'boxplot', fill = I('#D4A78E'),
      main = "Plot 1 - Mean Number of Years in School by Year") +
  ylab("Mean Number of Years in School")

qplot(data = filter(gather.school, year %in% c(1970,1980,1990,2000,2009)), 
      x = year, y = mean.num.years, 
      geom = 'boxplot', fill = I('#D4A78E'), 
      main = "Plot 2 - Mean Number of Years in School by 10 years intervals") +
  ylab("Mean Number of Years in School")

qplot(data = filter(gather.school, year %in% c(1970,1980,1990,2000,2009)), 
      x = mean.num.years, binwidth = .5, color = year, geom='freqpoly', 
      main = "Plot 3 - Countries' mean number of years in school by 10 years intervals") +
  scale_x_continuous(breaks = seq(0,15, 1)) +
  xlab("Mean Number of Years in School") +
  ylab("Number of countries")

```

The number of years spent in school has an overall increase through the years.  
In plot 3, we can observe that the number of countries with a value of 1 as mean number of years in school, is much lower in recent years. The opposite happens for a value of 12.  
However, looking at plot 1 and 2, one can also notice that the upper limits have been growing at a different rate than the lower limits.

***

### Range of values by decade
Now we take a closer look at an observation made before: the difference between the extremes in the distribution as well as the interquartile range of the values. The objective is to investigate the evolution of the gap between the countries with the highest and lowest values.

```{r}
diff.by.year <- gather.school %>% 
  group_by(year) %>% 
  arrange(mean.num.years) %>% 
  summarise(min = first(mean.num.years), max = last(mean.num.years))

diff.by.year %>% 
  mutate(decade = paste0(substr(year,1,3),0)) %>% 
  group_by(decade) %>% 
  summarise(min.value = mean(min), max.value = mean(max), diff.max.min = mean(max - min))
```

However, many times the extremes are not representative of the overall behavior of the distribution, so a better tool would be to use the interquartile range(IQR) to look into the range of the most common values.

```{r}
gather.school  %>% 
  group_by(year) %>% 
  summarise(iqr = IQR(mean.num.years)) %>% 
  mutate(decade = paste0(substr(year,1,3),0)) %>% 
  group_by(decade) %>% 
  summarise(mean.iqr = mean(iqr))
```

The gap between the countries with lower mean years spent in school and higher has been increasing per decade.

***

### Range of values by country
Let's look at the differences between 1970 and 2009 by country. That is, the difference between the mean number of years spent in school in 2009 and in 1970. This will tell us how much each country has changed in the past 4 decades, and in which direction.

```{r}
diff.1970.2009 <- mutate(gather.school %>% 
  group_by(country) %>% 
  summarise(in.1970 = first(mean.num.years), in.2009=last(mean.num.years)), diff = in.2009 - in.1970) 

summary(diff.1970.2009$diff)
```

The first observation is that every country has increased the mean number of years spent in school. Looking at the values of the mean and median we can also see that the average increase is just over 5 years, which represents a growth of more than 1 year per decade. We can plot these values to have a more detailed view.

```{r}
qplot(data = diff.1970.2009, x = diff, binwidth = 0.4, fill = I('light green'), color = I('black')) + 
  scale_x_continuous(breaks = seq(0.4, 9.5, 0.4)) +
  ylab("Number of countries") +
  xlab("Difference between 2009 and 1970 (mean number of years in school)")

```

#### Countries with the highest increase
```{r}
diff.1970.2009 %>% 
  arrange(desc(diff)) %>% 
  select(country, diff) %>% 
  slice(1:5)
```

#### Countries with lowest increase
```{r}
diff.1970.2009 %>% 
  arrange(diff) %>% 
  select(country, diff) %>% 
  slice(1:5)
```

***

### General Conclusions

With this study we are able to observe:  

1. The increase throughout the years, for every country, of the mean number of years spent in school.  
2. The range of values is increasing due to the bottom values increasing at a slower pace than the top values.  
3. The countries have been improving at very different paces.  

***

### Future Work
Gather more information on the countries that not only add more context to the study, but may also help understand some of the results that were presented before (highest and lowest values, ...).  
For example:

 1. Geographical information (continent, region);
 2. Cultural traditions;
 3. GDP of the countries;
 4. Important events (wars, natural disasters, new politics);

Also, one can choose an important event in a particular year and investigate if this event had any change in the values observed during and after. For example: a war (or the end of a war), a change in political regimes, etc..

***
Sara Carvalho 01-05-2015