Problem Set 3 - Facebook Friends
========================================================

This assignment is the last part of problem set 3 of the Udacity course *Data Analysis with R* lesson 3: *Explore One Variable*. This is a small task involving facebook friends' birthdays dataset.

## Dataset: Facebook Friends Birthdays

### Data Wrangling

```{r message=FALSE}
library(dplyr)
library(lubridate)
library(ggplot2)

bdays <- read.csv('birthdaysExample.csv')
str(bdays)
head(bdays)

bdays <- bdays %>% mutate(
  year = year(mdy(dates)),
  month = month(mdy(dates), label=TRUE), 
  n.month= month(mdy(dates)), 
  day = day(mdy(dates)), 
  wday = wday(mdy(dates), label=TRUE))

str(bdays)
head(bdays)
```

***

#### How many people share your birthday?
```{r}
# May 4th
bdays %>% 
  filter(day==4 & n.month==5) %>% 
  nrow() 
```

***

#### How many birthdays are there in each month? Which month contains the most birthdays?
```{r}
bdays %>% 
  group_by(month) %>% 
  summarise(n=n()) %>% 
  arrange(desc(n))
```

***

#### Which day has the most birthdays?
```{r}
bdays %>% 
  group_by(day) %>%
  summarise(n=n()) %>% 
  arrange(desc(n)) %>% 
  head()
```

***

#### Which day of the year has the most birthdays?
```{r}
bdays %>% 
  group_by(dates) %>% 
  summarise(day=first(day), month = first(month), n=n()) %>% arrange(desc(n))  %>% 
  select(day,month,n) %>% 
  head()
```

***

#### Which weekday has the most birthdays?
```{r}
bdays %>% 
  group_by(wday) %>% 
  summarise(n=n()) %>% 
  arrange(desc(n))
```

***

#### Do you have at least 365 friends that have birthdays on everyday of the year?
```{r fig.width=10}
qplot(x = day, data = bdays,binwidth = 1) + 
  scale_x_discrete(1:31) +
  facet_wrap(~month, ncol = 3)
```

By looking at the plot of the day distribution per month, no. Example: May, 3th...
Confirmation:
```{r}
bdays %>% 
  filter(day==3 & n.month==5) %>% 
  nrow() 
```

Number of days with no birthdays:
```{r}
365 - bdays %>% 
  group_by(month,day) %>% 
  summarise(n = n())  %>% 
  nrow()
```

***
Sara Carvalho
19-03-2015