Problem Set 4
========================================================

This assignment is the first part of problem set 4 of the Udacity course *Data Analysis with R* lesson 4: *Explore Two Variables*. This part of the problem set consists in a set of questions/tasks answered below.

### Diamonds
Load the diamonds dataset

```{r}
library(ggplot2)
data(diamonds)
str(diamonds)
```

***

### price vs. x (length in mm)
```{r}
ggplot(aes(x = x, y = price), data = diamonds) +
  geom_point()
```

#### Observations
Most of the diamonds have between 3 and 9 mm. The price increases rapidly with the value of x, in what seems to be a exponential relationship. There are some outliers of length 0, probably not the real value of length.

***

### Correlations

#### What is the correlation between price and x?
```{r}
cor.test(diamonds$x, diamonds$price)
```
0.88

#### What is the correlation between price and y?
```{r}
cor.test(diamonds$y, diamonds$price)
```
0.87

#### What is the correlation between price and z?
```{r}
cor.test(diamonds$z, diamonds$price)
```
0.86

***

###  price vs. depth (total depth percentage)
```{r}
ggplot(aes(x = depth , y = price), data = diamonds) + 
  geom_point()
```


####  Adjustments - price vs. depth
Plot with adjusted scale, based on the summary statistics, and points with alpha channel of 1/100.
```{r}
summary(diamonds$depth)

ggplot(aes(x = depth , y = price), data = diamonds) + 
  geom_point(alpha = 1/100) +
  scale_x_continuous(breaks = seq(43,80, 2))
```

#### Based on the scatter plot of depth vs. price, most diamonds are between what values of depth?
Between 59 - 64.

#### Correlation - price vs. depth
```{r}
cor.test(diamonds$depth, diamonds$price)
```
-0.01  
The value is close to 0 which indicates an extremely weak / nonexistent relationship between the two values.

#### Based on the correlation coefficient would you use depth to predict the price of diamonds?
No, the depth doesn't describe the price in a way that we can predict it's price based on a function (linear, exponential, etc..).

***

### price vs. carat
Create a scatter plot of price vs carat and omit the top 1% of price and carat values.

```{r}
ggplot(aes(x = carat, y = price), data = diamonds) +
  geom_point(alpha = 1/10) +
  xlim(0, quantile(diamonds$carat, 0.99)) +
  ylim(0, quantile(diamonds$price, 0.99))
```

***

### price vs. volume
Create a scatter plot of price vs. volume (x * y * z). This is a very rough approximation for a diamond's volume.  
Create a new variable for volume in the diamonds data frame.

```{r}
diamonds$volume <- diamonds$x * diamonds$y * diamonds$z

ggplot(aes(x = volume, y = price), data = diamonds) +
  geom_point()
```

#### Observations
Volume is mostly between 0+ and 500. The majority of prices seem to increase in an exponential way with volume.  
There are some outliers such as diamonds with 0 volume and diamonds with nearly 4000 in volume.
```{r warning=FALSE}
suppressMessages(library(plyr))
count(diamonds$volume == 0)
suppressMessages(detach("package:plyr", unload=TRUE))
```

***

### Correlations on Subsets
What's the correlation of price and volume?  
Exclude diamonds that have a volume of 0 or that are greater than or equal to 800.
```{r}
with(subset(diamonds, volume > 0 & volume < 800), cor.test(price, volume))
```
0.92  
This value indicates a strong positive relationship between the two values.

***

### Adjustments - price vs. volume
Subset the data to exclude diamonds with a volume greater than or equal to 800. Also, exclude diamonds with a volume of 0. Adjust the transparency of the points and add a linear model to the plot.
```{r}
ggplot(aes(x = volume, y = price), 
       data = subset(diamonds, volume > 0 & volume < 800)) +
  geom_point(alpha = 1/5, color = 'orange') +
  geom_smooth(method = 'lm', color = 'red') +
  geom_smooth(method = 'lm', color = 'green', formula = y ~ poly(x,2, raw = TRUE)) +
  geom_smooth(method = 'lm', color = 'blue', formula = y ~ poly(x,3, raw = TRUE)) +
  geom_smooth(method = 'lm', color = 'black', formula = y ~poly(x,4,raw=TRUE)) + 
  coord_cartesian(xlim = c(0,600), ylim=c(0,20000))
```

Red, linear model, fits the data reasonably well and is the simplest method in the graph, therefore not very expensive to use.  
The green model, $x^{2}$, blue model, $x^{3}$, and black model, $x^{4}$, seem to adapt, respectively, more accurately to the data.

***

### Mean Price by Clarity
Use the function *summarise* from the *dplyr* package to create a new data frame containing info on diamonds by clarity.  
Add the variables *mean_price*, *median_price*, *min_price*, *max_price* and *n*.
```{r}
suppressMessages(library(dplyr))
diamonds.by.clarity <- diamonds %>%
  group_by(clarity) %>%
  summarise(min_price = min(price),
            median_price = median(as.numeric(price)),
            mean_price = mean(price),
            max_price = max(price),
            n = n()) 

diamonds.by.clarity
```

***

###  Bar Charts of Mean Price
We’ve created summary data frames with the mean price by clarity and color. You can run the code in R to verify what data is in the variables *diamonds_mp_by_clarity* and *diamonds_mp_by_color*.  
Your task is to write additional code to create two bar plots on one output image using the *grid.arrange()* function from the package *gridExtra*.

```{r}
diamonds_by_clarity <- group_by(diamonds, clarity)
diamonds_mp_by_clarity <- summarise(diamonds_by_clarity, mean_price = mean(price))

diamonds_by_color <- group_by(diamonds, color)
diamonds_mp_by_color <- summarise(diamonds_by_color, mean_price = mean(price))

suppressMessages(library(gridExtra))

p1 <- ggplot(aes(x = clarity, y = mean_price), data = diamonds_mp_by_clarity) +
  geom_bar(stat = 'identity', fill  = 'orange') +
  scale_y_continuous(breaks=seq(0,5200, 500)) +
  coord_flip() 

p2 <- ggplot(aes(x = color, y = mean_price), data = diamonds_mp_by_color) +
  geom_bar(stat = 'identity', fill = 'light blue') +
  scale_y_continuous(breaks=seq(0,5200, 500)) +
  coord_flip()
  
grid.arrange(p1,p2)
```

#### Observations
Higher mean prices for color J and for clarity SI2 (both over 5000). Lower mean prices for clarity VVS1 (around 2500) and for color E and D (around 3000).  
Mean prices for color look odd since the prices are decreasing as the color type is improving. This can also be observed between some clarity levels.

***
Sara Carvalho
05-03-2015